Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-06-27 draft.tex
Included revised version from Steffi

Cohort studies comprise of heterogenous variables describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.

@@

\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=2.8in]{figures/workflow_comparison}
 \caption{
 (a) The standard epidemiology workflow consists of four steps.
 %
 (b) \emph{IVA} tools complement parts of this workflow instead of replacing them.
 %
 The combination of statistical and interactive analysis shows promising potential to unveil information in the data.
 %
 %IVA complements this workflow and is used to elaborate the feature selection.
 %
 We call the iterative red highlighted part \emph{IVA Loop}, described in detail in Figure~\ref{fig:InteractionLoop}.
 %(a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.
 %
 }
  \label{fig:WorkflowComparison}
\end{figure}
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics, and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew et al. \cite{Thew2009}, the workflow can be characterized as follows:
%
\begin{enumerate}\itemsep0.1em
	\item A hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of variables depicting conditions affected by the hypothesis is compiled accordingly.
%
	\item Confounding variables are identified and taken into account (for example using stratification).
%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach. %although it does not consider image data.
%
We focus, on the potential of image data and attempt to support \emph{hypothesis generation}.

\emph{Reproducibility} of results is an epidemiological key requirement.
%
It is difficult to achieve, since many physicians are involved when thousands of subjects are examined and interviewed.
%
Thus, both intra- and inter-observer variability needs to be low for all aspects of a cohort study examination.
%
Longitudinal studies require the acquired variables to be comparable for evaluation.
%
%If the data acquisition process changes, an information bias is introduced to the data, hampering inference in and between acquisition cycles.
%
%
Grouping subjects using epidemiological variables is essential in cohort studies to allow per-group risk determination.
%
Grouping depends on the underlying hypothesis.
%
Age for example is divided into groups (e.g. in 20 year steps) when investigating its influence.
%
These groups strongly depend on the condition of interest and therefore there is no standard for their categorization.
% --- NEW PARAGRAPH

@@

Indicators for medical conditions as well as questions about a subject's lifestyle are often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
%
Dichotomous data can also be derived by aggregating variables to yield only two manifestations (e.g. subjects younger or older than 50 years).
%
Medical examinations comprise categorical (e.g. levels of back pain) and continuous values (e.g. age or body size).
%
Data analysis is usually carried out by calculating correlations,
which is challenging due to the data type heterogeneity.
%
Parameter correlation can also be associated with confounding, which can not be automatically predicted. 
%
It has to be judged by a domain expert.
%
Sparse populated variables are hard to assess statistically.
%
Too few data samples may distort the real underlying distributions.
%
Statistical correlations are prone to \emph{confounding}, meaning that the association of two variables is influenced by a third variable, which needs to be isolated.
%
A famous example is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation. %than people with small shoe size.
%
The shoe size is actually associated with gender, where women have smaller feet and a longer life expectation.
%Also, correlation can both be of epidemiological interest but

@@

Steenwijk et al. \cite{Steenwijk2010} propose a relational database to organize cohort study data for a visual analysis based on linked views such as parallel coordinates, scatterplots and time plots.
%
Information about medical image data is incorporated via mappers, which extract comparable metrics about the data.
%
Medical image data can be displayed individually for subjects, e.g., for analyzing outliers.
%
While we use a similar approach when analyzing non-image data, our process also includes \emph{overview visualizations} and statistical suggestions of potentially interesting variables.
%

@@

Angelelli et al. \cite{Angelelli2014} focus on a data organization for the interactive visual analysis of heterogeneous cohort study data.
%
%\rem{They propose a data-cube based model able to handle partially overlapping data subsets during the interactive visualization.}
%
Brain image data was integrated into the analysis by first, segmenting brain regions and tracking neural pathways and then, deriving attributes from both, e.g., volume and fractional anisotropy.
%
The data-cube model facilitates the seamless integration of heterogeneous image-based and non-image data.
%
A multiple coordinated view framework linked spatial and non-spatial data views.
%
%The approach is a demonstrated by a case-study analysis of selected aspects of brain connectivity.
% --- Steffens Fazit ---
%
Our integration of image data into the analysis is similar to the work of Steenwijk \cite{Steenwijk2010} and Angelelli \cite{Angelelli2014} and colleagues.
%
While they offer a single spatial view for visualizing image-based information of one subgroup of the cohort, we provide multiple views showing the information of subgroups and their respective deviation from the entire cohort.
%
%\rem{We also integrate the data-driven definition of subgroups using clustering, statistically-based suggestions of potentially interesting features, and adaptive feature visualization.} %tailored to epidemiological data, e.g., a mosaic plot for the frequent categorical variables.
% %
% %Another unique characteristic is our web-based implementation and the integration of D3.js.
% %
% %The latter offers an extensive set of information visualization plots, which are at the disposal of our system.
% %
% --- / Steffens Fazit ---

@@

Due to the short paper length, detail is missing on the data type and their algorithm of identifying similar subjects works or what and if they employ statistical measures.

@@

Histograms, bar charts, scatterplots and heat maps are used to visualize variable combinations with regard to their type.

@@

Dai et al. \cite{Dai2005} explored risk factors by incorporating choropleth maps of epidemiological variables (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines.

@@

While their goal to identify possible risk factors using socio-economic and health data is similar to ours, they focus on iteratively refining defined hypothesis and on geographical data.

@@

While the work shows, how different visualization techniques provide insight into these data sets, it focuses on the time aspect, which is not present for our data.

@@

% *********************** Our own Work ***********************
% Prior Work
\paragraph{Prior Work.}
Klemm et al. \cite{Klemm2013VMV} visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort.
%
Hierarchical agglomerative clustering divided the population into shape-related groups.
%
As proof of concept, a relation between the size of the segmented shape and measured size of the subjects was shown.
%
This work focuses on incorporating these derived data as new variables, making it possible to include it into the hypothesis validation and generation process.
%
When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
%
Niemann et al. \cite{Niemann2014} presented an interactive data mining tool for the assessment of risk factors of hepatic steatosis, the fatty liver disease.
%
Association rules created by data mining methods can be analyzed interactively with their tool and highlight potentially overlooked variables.
%
%\add{We combine the idea of suggesting potentially interesting features by adding them to the \emph{IVA}-workflow in order to amplify hypothesis generation.}
% *********************** / Our own Work ***********************

@@

Lammarsch et al. \cite{Lammarsch2011} provided a workflow and terminology definition of Visual Analysis techniques.
%
They define a model as a representation of system entities, phenomena and processes and hypothesis as models whose outcomes are not compared with real-world data (\emph{validation}).
%
The VA-process is also reflected in our \emph{IVA} loop.
% --- PARAGRAPH ---

Baldonado et al. \cite{Baldonado2000} presented rules for designing multiple coordinated views.
%
They point out the cost-benefit-tradeoff introduced by the cognitive overhead by mentally connecting multiple views over more complex single views. %introduced by tasks, such as view comparison or context switching.
%
Weaver et al. \cite{Weaver2010} extracted guidelines for cross-filtering multiple views by incorporating \emph{views} mapping data to visual elements, \emph{brushes} for selecting these elements and \emph{switches} for linking brushing results between \emph{views}.
%
Our system follows the same rules for selecting subject groups, but our goal is to judge variable relations and potential outcomes.
\\\\
%Related work focused on epidemiological data often deals with time-dependent data, hypothesis generation by guidance through parameter suggestions is seldom.
%The uniqueness of the \emph{IVA} workflow applied to image-based cohort study data compared to the discussed work is threefold.
The uniqueness of our workflow compared to the discussed work is threefold.
%
(1) We incorporate 3D models abstracting shape information fused with non-image data visualizations, allowing to analyze local physiological changes related to non-image parameters.
%
(2) We focus on hypothesis generation by discovering new relationships associated with shape information. %, as epidemiologists already have an well established workflow for statistically validating hypothesis.
%
(3) Overview visualizations using statistical abstractions aim to provide an unbiased variable relationship assessment.

%\section{Interactive Visual Analyis in Cohort Study Data}
\section{Image Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
We described the epidemiological workflow and emphasized the reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
%
%Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as a consecutive series of steps.
%
%Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate its weaknesses rather than replacing the existing workflow.
Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate the weaknesses of the existing workflow, rather than replacing it (recall Fig.~\ref{fig:WorkflowComparison}).
%
In the current state, the workflow treats the data like a black box.
%
Statistical tests on variables associated to a hypothesis yield a value for deciding whether the data supports the hypothesis.
%
Variables not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted.
%
This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
domain experts annotate landmarks tediously to derive measures, such as diameters.
%
This leaves out the majority of information in the image data by abstracting it to single values.
%
%It is easily possible that omitted information would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
\\\\
\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative variable selection process (see Fig.~\ref{fig:WorkflowComparison}~b).
%\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
%
It also aims to project back into the hypothesis formulation step to amplify hypothesis generation.
%
This has to be handled with care, since \emph{overfitting} of expectations to the data is an imminent danger \cite{Turkay2013}.

@@

\paragraph{Multivariate Analysis} refers to an investigation of multi-variate properties of the dependent data by specifying a variable in one attribute view while analyzing the value distribution with respect to other variables in other attribute views.
%
Epidemiologists may define a variable in a scatter plot of the body mass index (BMI) and age to inspect the result in a histogram of body height.
%
These associations may also be summarized using pivot tables which are widely used in epidemiology.

@@

This is either done manually by experts setting landmarks or by a (semi-)automatic detection, registration and segmentation.

@@

Morphometric variables are derived to allow for statistical comparison of the tissue, which incorporates mostly positions, diameters, volumes and relative distances and alignment to other structures.

@@

 The data are visualized in the range and domain space, which can be brushed, yielding new groups, to be investigated using further variables.

@@

 Distance to the mean mesh of subjects with thyroid disorder is encoded as red for x axis, blue to y axis and green to z axis.

@@

Our proposed \emph{IVA} workflow consists of three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Variable selection, visualization and brushing.

@@

A hypothesis-driven analysis usually starts with the selection of variables, or shape groups derived from a shape-based clustering.

@@

% \subsection{Interaction and Visualization Techniques} \label{Interaction- and Visualization Techniques}
\section{System Design and Implementation} \label{Interaction- and Visualization Techniques}
%
The suitability of visualization techniques for epidemiological data depends on its ability to compare multiple data variables while highlighting associations.
%
%The methods have to reflect the routines that epidemiologists take in order to be incorporated into their research.
%
Visual evaluations of data are therefore as important as methods allowing for numerical data analysis.
%
In the following sections we present the different parts of system.
% \subsection{System Structure and Design} \label{Structure and Workflow}
\subsection{Design and Visualization Techniques} \label{Structure and Workflow}
% \begin{itemize}
% 	\item [x] Was sind die requirements des tools?
% 	\item [x] how it was decided to create this particular design
% 	\item [x] which kind of User interaction was made?
% 	\item [x] Spezielle Anforderungen durch Webseitenformat, weggehen von allgemeinen UI Elementen wie Menubar, keine Rechtsklickmenüs.
% 	\item [x] Entschieden eine saubere UI zu bauen, Unterscheide durch Whitespace akzentuieren. Interface in Fig. 3 zu sehen.
% 	\item [x] Wie wird interagiert, wie wird gebrusht, was wird gelinkt?
% 	\item [x] wie wird das Clustering initialisiert?
% 	\item auf Kontingenzmatrix mit verweisen als Teil des Interfaces (dieses auch noch anpassen)
% \end{itemize}
%Early it became clear that we have to rely a lot on online communication with our clinical partners due to their busy schedule and our large spatial distance towards each other.
Early it became clear that we have to rely a lot on online communication due to the large spatial distance towards each other.
%
Hence we build our system using web technologies.%, described in detail in Section~\ref{implementation}.
%
By running the prototypes on server machines, software exchange became as easy as sharing a website link, giving us the opportunity to include the clinical experts in the development process with little effort.
%
Incorporating the \emph{IVA} workflow for image-centric cohort study data requires \emph{overview visualizations} as well as \emph{multivariate visualizations}, which bring image-derived information in context to non-image variables.

The focus on web technologies is not without tradeoffs.
%
Classical UI elements, such as the menu bar or custom right click menus, are technically possible, but not common in this domain.
%
In favor of a clean layout, we designed the system without such components.
%
Since the previously described \emph{IVA} workflow allows for many different ways to analyze the data, we tried to make the interface as minimalistic as possible, treating the resulting space as \emph{canvas} for the data.
%
% \subsubsection{System Structure}
We divide the workspace into four major parts, as illustrated in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
\begin{itemize} \itemsep0.2em
		\item The \emph{sidebar}, which contains all epidemiological variables. As cluster results group variables like categorical variables and are part of the sidebar as well (Fig.~\ref{fig:visualization}~a).
	%
	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely (Fig.~\ref{fig:visualization}~b).
	%
	\item The interactive \emph{pivot table} gives detailed numerical information of the variables in the canvas view. This view on the data is familiar to epidemiologists (Fig.~\ref{fig:visualization}~c).
	%
	\item The \emph{contingency view} depicts relations for variables in the canvas in an contingency matrix (Fig.~\ref{fig:similarity}) and a \emph{contingency list}.
\end{itemize}

\paragraph{System Layout.}
We experimented with several layouts.
%
The initial idea was to make all components freely arrange- and resizable on a large \emph{canvas} area.
%
This idea was soon dropped since domain experts reported a cluttered workspace, which required a lot of scrolling.
%
The introduction of separate panes for the contingency matrix, pivot table and sidebar, displayed with a mouse click on the corresponding button and sliding on top of the \emph{canvas} was considered more feasible (Figure~\ref{fig:visualization} shows the system with reeled-out pivot table pane).
%
All user-generated visualizations follow are part of the \emph{canvas} and can be arranged freely.
% are the  was to keep the concept of free arrangement for user-added 
%The resulting system can be seen in Figure~\ref{fig:visualization} (left).

@@

\paragraph{Adaptive Variable Visualization.} \label{sec:AdaptiveFeatureVisualization}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/similarity_matrix}
 \caption{Contingency matrix of 129 variables (127 data set variables, 2 cluster results) showing 16,641 combinations.
 %
 Similarity is calculated using the \emph{Cram\'{e}r's V} contingency value.
 %
 Color brightness encodes association strength.
 %Mouse-over focuses a entry and highlights the feature names for better readability.
 Mouse-over an entry enlarges the variable names for better readability.
 %
 The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
 }
 \label{fig:similarity}
\end{figure}
The visualization type, inspired by \texttt{GPLOMS} \cite{GPLOMS, Francois2013}, is dynamically chosen based on the variable types and number to allow for \emph{multivariate analysis}.
%
Categorical data are either mapped to bar charts (single variables) or mosaic plots (multiple variables).
%
Figure \ref{fig:visualization} describes this dynamic adjustment.
%
Continuous data can be visualized using scatterplots (two variables) or parallel coordinates (multiple variables), but in epidemiology, this data type is usually categorized into ordinal groups of \emph{equal size}.
%
Since the number of categories often depends on the hypothesis, the discretization steps can be adapted dynamically.
%
Too many groups potentially generate sparse bins not suited for statistical evaluation.
%
Not enough groups overgeneralize information.
%
Adaptive discretization is an option, but imposes possible overfitting to the data.
%
Conclusions based on statistical relationships derived from groups already biased by variable distribution are heavily influenced by the used discretization.
%
Therefore we follow the convention to use bins of equal size.

>>>>

Cohort studies comprise of heterogenous variables describing the medical condition as well as demographic and lifestyle factors and, more recently, medical image data.

@@

\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=2.8in]{figures/workflow_comparison}
 \caption{
 (a) The standard epidemiology workflow consists of four steps.
 %
 (b) \emph{IVA} tools complement parts of this workflow instead of replacing them.
 %
 The combination of statistical and interactive analysis shows promising potential to unveil information in the data.
 %
 %IVA complements this workflow and is used to elaborate the feature selection.
 %
 We call the iterative red highlighted part \emph{IVA Loop}, described in detail in Figure~\ref{fig:InteractionLoop}.
 %(a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.
 %
 }
  \label{fig:WorkflowComparison}
\end{figure}
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics, and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew et al. \cite{Thew2009}, the workflow can be characterized as follows:
%
\begin{enumerate}\itemsep0.1em
	\item A hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of variables depicting conditions affected by the hypothesis is compiled accordingly.
%
	\item Confounding variables are identified and taken into account (for example using stratification).
%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach. %although it does not consider image data.
%
We focus on the potential of image data and attempt to support \emph{hypothesis generation}.

\emph{Reproducibility} of results is an epidemiological key requirement.
%
It is difficult to achieve, since many physicians are involved when thousands of subjects are examined and interviewed.
%
Thus, both intra- and inter-observer variability needs to be low for all aspects of a cohort study examination.
%
Longitudinal studies require the acquired variables to be comparable for evaluation.
%
%If the data acquisition process changes, an information bias is introduced to the data, hampering inference in and between acquisition cycles.
%
%
Grouping subjects using epidemiological variables is essential in cohort studies to allow per-group risk determination.
%
Grouping depends on the underlying hypothesis.
%
Age, for example, is divided into groups (e.g. in 20 year steps) when investigating its influence.
%
These groups strongly depend on the condition of interest and therefore there is no standard for their categorization.
% --- NEW PARAGRAPH

@@

Indicators for medical conditions as well as questions about a subject's lifestyle are often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
%
Dichotomous data can also be derived by aggregating variables to yield only two manifestations (e.g. subjects younger or older than 50 years).
%
Medical examinations comprise categorical (e.g. levels of back pain) and continuous values (e.g. age or body size).
%
Data analysis is usually carried out by calculating correlations,
which is challenging due to the data type heterogeneity.
%
Parameter correlation can also be associated with confounding, which cannot be automatically predicted. 
%
It has to be judged by a domain expert.
%
Sparse populated variables are hard to assess statistically.
%
Too few data samples may distort the real underlying distributions.
%
Statistical correlations are prone to \emph{confounding}, meaning that the association of two variables is influenced by a third variable, which needs to be isolated.
%
A famous example is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation. %than people with small shoe size.
%
The shoe size is actually associated with gender, where women have smaller feet and a longer life expectation.
%Also, correlation can both be of epidemiological interest but

@@

Steenwijk et al. \cite{Steenwijk2010} propose a relational database to organize cohort study data for a visual analysis based on linked views such as parallel coordinates, scatter plots and time plots.
%
Information about medical image data is incorporated via mappers, which extract comparable metrics about the data.
%
Medical image data can be displayed individually for subjects, e.g., for analyzing outliers.
%
While we use a similar approach when analyzing non-image data, our process also includes \emph{overview visualizations} and statistical suggestions of potentially interesting variables.
%

@@

Angelelli et al. \cite{Angelelli2014} focus on a data organization for the interactive visual analysis of heterogeneous cohort study data.
%
%\rem{They propose a data-cube based model able to handle partially overlapping data subsets during the interactive visualization.}
%
Brain image data was integrated into the analysis by first segmenting brain regions and tracking neural pathways and then deriving attributes from both, e.g., volume and fractional anisotropy.
%
The data-cube model facilitates the seamless integration of heterogeneous image-based and non-image data.
%
A multiple coordinated view framework linked spatial and non-spatial data views.
%
%The approach is a demonstrated by a case-study analysis of selected aspects of brain connectivity.
% --- Steffens Fazit ---
%
Our integration of image data into the analysis is similar to the work of Steenwijk \cite{Steenwijk2010} and Angelelli \cite{Angelelli2014} and colleagues.
%
While they offer a single spatial view for visualizing image-based information of one subgroup of the cohort, we provide multiple views showing the information of subgroups and their respective deviation from the entire cohort.
%
%\rem{We also integrate the data-driven definition of subgroups using clustering, statistically-based suggestions of potentially interesting features, and adaptive feature visualization.} %tailored to epidemiological data, e.g., a mosaic plot for the frequent categorical variables.
% %
% %Another unique characteristic is our web-based implementation and the integration of D3.js.
% %
% %The latter offers an extensive set of information visualization plots, which are at the disposal of our system.
% %
% --- / Steffens Fazit ---

@@

Due to the short paper length, detail is missing on the data type and their algorithm of identifying similar subjects works or whether they employ statistical measures.

@@

Histograms, bar charts, scatter plots and heat maps are used to visualize variable combinations with regard to their type.

@@

Dai et al. \cite{Dai2005} explored risk factors by incorporating choropleth maps of epidemiological variables (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatter plots with integrated regression lines.

@@

While their goal to identify possible risk factors using socio-economic and health data is similar to ours, they focus on iteratively refining defined hypotheses and on geographical data.

@@

While the work shows how different visualization techniques provide insight into these data sets, it focuses on the time aspect, which is not present for our data.

@@

% *********************** Our own Work ***********************
% Prior Work
\paragraph{Prior Work.}
We \cite{Klemm2013VMV} visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort.
%
Hierarchical agglomerative clustering divided the population into shape-related groups.
%
As proof of concept, a relation between the size of the segmented shape and the measured size of the subjects was shown.
%
This work focuses on incorporating these derived data as new variables, enabling to include it into the hypothesis validation and generation process.
%
When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
%
Niemann et al. \cite{Niemann2014} presented an interactive data mining tool for the assessment of risk factors of hepatic steatosis, the fatty liver disease.
%
Association rules created by data mining methods can be analyzed interactively with their tool and highlight potentially overlooked variables.
%
%\add{We combine the idea of suggesting potentially interesting features by adding them to the \emph{IVA}-workflow in order to amplify hypothesis generation.}
% *********************** / Our own Work ***********************

@@

Lammarsch et al. \cite{Lammarsch2011} provided a workflow and terminology definition of Visual Analysis techniques.
%
They define a model as a representation of system entities, phenomena, processes and hypotheses as models whose outcomes are not compared with real-world data (\emph{validation}).
%
The VA process is also reflected in our \emph{IVA} loop.
% --- PARAGRAPH ---

Baldonado et al. \cite{Baldonado2000} presented rules for designing multiple coordinated views.
%
They point out the cost-benefit tradeoff introduced by the cognitive overhead by mentally connecting multiple views over more complex single views. %introduced by tasks, such as view comparison or context switching.
%
Weaver et al. \cite{Weaver2010} extracted guidelines for cross-filtering multiple views by incorporating \emph{views} mapping data to visual elements, \emph{brushes} for selecting these elements and \emph{switches} for linking brushing results between \emph{views}.
%
Our system follows the same rules for selecting subject groups, but our goal is to judge variable relations and potential outcomes.
\\\\
%Related work focused on epidemiological data often deals with time-dependent data, hypothesis generation by guidance through parameter suggestions is seldom.
%The uniqueness of the \emph{IVA} workflow applied to image-based cohort study data compared to the discussed work is threefold.
The uniqueness of our workflow compared to the discussed work is threefold.
%
(1) We incorporate 3D models abstracting shape information fused with non-image data visualizations, allowing to analyze local physiological changes related to non-image parameters.
%
(2) We focus on hypothesis generation by discovering new relationships associated with shape information. %, as epidemiologists already have an well established workflow for statistically validating hypothesis.
%
(3) Overview visualizations using statistical abstractions aim to provide an unbiased variable relationship assessment.

%\section{Interactive Visual Analyis in Cohort Study Data}
\section{Image-Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
We described the epidemiological workflow and emphasized the reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
%
%Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as a consecutive series of steps.
%
%Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate its weaknesses rather than replacing the existing workflow.
Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate the weaknesses of the existing workflow rather than replacing it (recall Fig.~\ref{fig:WorkflowComparison}).
%
In the current state, the workflow treats the data like a black box.
%
Statistical tests on variables associated to a hypothesis yield a value for deciding whether the data supports the hypothesis.
%
Variables not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted.
%
This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
domain experts annotate landmarks tediously to derive measures, such as diameters.
%
This leaves out the majority of information in the image data by abstracting it to single values.
%
%It is easily possible that omitted information would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and could also identify possible anatomical confounders--an epidemiolgical research result in itself.
\\\\
\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative variable selection process (see Fig.~\ref{fig:WorkflowComparison}~b).
%\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
%
It also aims to project back into the hypothesis formulation step to amplify hypothesis generation.
%
This has to be handled with care, since \emph{overfitting} of expectations to the data is an imminent danger \cite{Turkay2013}.

@@

\paragraph{Multivariate Analysis} refers to an investigation of multi-variate properties of the dependent data by specifying a variable in one attribute view while analyzing the value distribution with respect to other variables in other attribute views.
%
Epidemiologists may define a variable in a scatter plot of the body mass index (BMI) and age to inspect the result in a histogram of body height.
%
These associations may also be summarized using pivot tables, which are widely used in epidemiology.

@@

This is either done manually by experts setting, landmarks or by a (semi-)automatic detection, registration and segmentation.

@@

Morphometric variables are derived to allow for statistical comparison of the tissue, which incorporates mostly positions, diameters, volumes, relative distances and alignment to other structures.

@@

 The data are visualized in the range and domain space, which can be brushed, yielding new groups to be investigated using further variables.

@@

 Distance to the mean mesh of subjects with thyroid disorder is encoded as red for x axis, blue y axis and green z axis.

@@

Our proposed \emph{IVA} workflow consists of three major steps, as illustrated in Figure~\ref{fig:InteractionLoop}: Variable selection, visualization and brushing.

@@

A hypothesis-driven analysis usually starts with the selection of variables or shape groups derived from a shape-based clustering.

@@

% \subsection{Interaction and Visualization Techniques} \label{Interaction- and Visualization Techniques}
\section{System Design and Implementation} \label{Interaction- and Visualization Techniques}
%
The suitability of visualization techniques for epidemiological data depends on its ability to compare multiple data variables while highlighting associations.
%
%The methods have to reflect the routines that epidemiologists take in order to be incorporated into their research.
%
Visual evaluations of data are therefore as important as methods allowing for numerical data analysis.
%
In the following sections we present the different parts of the system.
% \subsection{System Structure and Design} \label{Structure and Workflow}
\subsection{Design and Visualization Techniques} \label{Structure and Workflow}
% \begin{itemize}
% 	\item [x] Was sind die requirements des tools?
% 	\item [x] how it was decided to create this particular design
% 	\item [x] which kind of User interaction was made?
% 	\item [x] Spezielle Anforderungen durch Webseitenformat, weggehen von allgemeinen UI Elementen wie Menubar, keine Rechtsklickmenüs.
% 	\item [x] Entschieden eine saubere UI zu bauen, Unterscheide durch Whitespace akzentuieren. Interface in Fig. 3 zu sehen.
% 	\item [x] Wie wird interagiert, wie wird gebrusht, was wird gelinkt?
% 	\item [x] wie wird das Clustering initialisiert?
% 	\item auf Kontingenzmatrix mit verweisen als Teil des Interfaces (dieses auch noch anpassen)
% \end{itemize}
%Early it became clear that we have to rely a lot on online communication with our clinical partners due to their busy schedule and our large spatial distance towards each other.
Early it became clear that we have to rely a lot on online communication due to the large spatial distance towards each other.
%
Hence, we build our system using web technologies. %, described in detail in Section~\ref{implementation}.
%
By running the prototypes on server machines, software exchange became as easy as sharing a website link, giving us the opportunity to include the clinical experts in the development process with little effort.
%
Incorporating the \emph{IVA} workflow for image-centric cohort study data requires \emph{overview visualizations} as well as \emph{multivariate visualizations}, which bring image-derived information in context to non-image variables.

The focus on web technologies is not without tradeoffs.
%
Classical UI elements, such as the menu bar or custom right-click menus, are technically possible, but not common in this domain.
%
In favor of a clean layout, we designed the system without such components.
%
Since the previously described \emph{IVA} workflow allows for many different ways to analyze the data, we tried to make the interface as minimalistic as possible, treating the resulting space as \emph{canvas} for the data.
%
% \subsubsection{System Structure}
We divide the workspace into four major parts, as illustrated in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
\begin{itemize} \itemsep0.2em
		\item The \emph{sidebar}, which contains all epidemiological variables. The cluster results group variables like categorical variables and are part of the sidebar as well (Fig.~\ref{fig:visualization}~a).
	%
	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely (Fig.~\ref{fig:visualization}~b).
	%
	\item The interactive \emph{pivot table} gives detailed numerical information of the variables in the canvas view. This view on the data is familiar to epidemiologists (Fig.~\ref{fig:visualization}~c).
	%
	\item The \emph{contingency view} depicts relations for variables in the canvas in an contingency matrix (Fig.~\ref{fig:similarity}) and a \emph{contingency list}.
\end{itemize}

\paragraph{System Layout.}
We experimented with several layouts.
%
The initial idea was to make all components freely arrange- and resizable on a large \emph{canvas} area.
%
This idea was soon dropped, since domain experts reported a cluttered workspace, which required a lot of scrolling.
%
The introduction of separate panes for the contingency matrix, pivot table and sidebar, displayed with a mouse click on the corresponding button and sliding on top of the \emph{canvas} was considered more feasible (Figure~\ref{fig:visualization} shows the system with reeled-out pivot table pane).
%
All user-generated visualizations are part of the \emph{canvas} and can be arranged freely.
% are the  was to keep the concept of free arrangement for user-added 
%The resulting system can be seen in Figure~\ref{fig:visualization} (left).

@@

\paragraph{Adaptive Variable Visualization.} \label{sec:AdaptiveFeatureVisualization}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/similarity_matrix}
 \caption{Contingency matrix of 129 variables (127 data set variables, 2 cluster results) showing 16,641 combinations.
 %
 Similarity is calculated using the \emph{Cram\'{e}r's V} contingency value.
 %
 Color brightness encodes association strength.
 %Mouse-over focuses a entry and highlights the feature names for better readability.
 Moving the mouse over an entry enlarges the variable names for better readability.
 % Potential Cut here
 The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
 }
 \label{fig:similarity}
\end{figure}
The visualization type, inspired by \texttt{GPLOMS} \cite{GPLOMS, Francois2013}, is dynamically chosen based on the variable types and number to allow for \emph{multivariate analysis}.
%
Categorical data are either mapped to bar charts (single variables) or mosaic plots (multiple variables).
%
Figure \ref{fig:visualization} describes this dynamic adjustment.
%
Continuous data can be visualized using scatter plots (two variables) or parallel coordinates (multiple variables), but in epidemiology, this data type is usually categorized into ordinal groups of \emph{equal size}.
%
Since the number of categories often depends on the hypothesis, the discretization steps can be adapted dynamically.
%
Too many groups potentially generate sparse bins not suited for statistical evaluation.
%
Not enough groups overgeneralize information.
%
Adaptive discretization is an option, but imposes possible overfitting to the data.
%
Conclusions based on statistical relationships derived from groups already biased by variable distribution are heavily influenced by the used discretization.
%
Therefore, we follow the convention to use bins of equal size.
%
